Monte Carlo Expectation Maximization for RNAseq

### 2015-06-22 slides/EM/em.Rnw
+dplyr

<<chunk_options, echo=FALSE, message=FALSE>>=
opts_chunk$set(fig.width=6, fig.height=5, out.width='.8\\linewidth', fig.align='center', size='tiny')
library(plyr)
library(ggplot2)
library(reshape2)
library(rstan)
@

@@

>>>>

<<chunk_options, echo=FALSE, message=FALSE>>=
opts_chunk$set(fig.width=6, fig.height=5, out.width='.8\\linewidth', fig.align='center', size='tiny')
library(plyr)
library(dplyr)
library(ggplot2)
library(reshape2)
library(rstan)
@

@@

\section{Heterosis}
\begin{frame}[fragile]
\frametitle{Heterosis}

<<heterosis_model, cache=TRUE>>=
heterosis_model = "
data {
  int<lower=1> n;
  int<lower=0> count[n];
  int<lower=1, upper=3> variety[n];
  real mu_phi;
  real mu_alpha;
  real mu_delta;
  real mu_psi;
  real<lower=0> tau_phi;
  real<lower=0> tau_alpha;
  real<lower=0> tau_delta;
  real<lower=0> tau_psi;
}
transformed data {
  matrix[n,3] X;
  for (i in 1:n) {
    if (variety[i] == 1) { X[i,1] <-  1; X[i,2] <- -1; X[i,3] <-  0; }
    if (variety[i] == 2) { X[i,1] <-  1; X[i,2] <-  1; X[i,3] <-  0; }
    if (variety[i] == 3) { X[i,1] <-  1; X[i,2] <-  0; X[i,3] <-  1; }
  }
}
parameters {
  real phi;
  real alpha;
  real delta;
  real psi;
}
transformed parameters {
  vector[3] pad;

  pad[1] <- phi;
  pad[2] <- alpha;
  pad[3] <- delta;
}
model {
  phi   ~ normal(mu_phi  , sqrt(tau_phi  ));
  alpha ~ normal(mu_alpha, sqrt(tau_alpha));
  delta ~ normal(mu_delta, sqrt(tau_delta));
  psi   ~ normal(mu_psi  , sqrt(tau_psi  ));
  count ~ neg_binomial_2_log(X*pad, 1/exp(psi));
}
"
h_m = stan_model(model_code = heterosis_model)
@
\end{frame}

\begin{frame}[fragile]
<<heterosis_data, cache=TRUE>>=
G = 100
n = 4
truth = data.frame(mu_phi   = 3, tau_phi   = 1, mu_psi = 0, tau_psi = 1,
                   mu_alpha = 0, tau_alpha = 1, mu_delta = 0, tau_delta = .1^2)
par = data.frame(gene=1:G, 
                 phi   = with(truth, rnorm(G, mu_phi  , sqrt(tau_phi  ))),
                 alpha = with(truth, rnorm(G, mu_alpha, sqrt(tau_alpha))),
                 delta = with(truth, rnorm(G, mu_delta, sqrt(tau_delta))),
                 psi   = with(truth, rnorm(G, mu_psi  , sqrt(tau_psi  ))))
X = cbind(1, rep(c(-1,1,0), each=n), rep(c(0,0,1), each=n))
d = ddply(par, .(gene), function(x) {
  pad = with(x, c(phi,alpha,delta))
  data.frame(y = rnbinom(3*n, mu = exp(X%*%pad), size=1/exp(x$psi)),
             variety = rep(1:3, each=n))
})
@
\end{frame}

\begin{frame}[fragile]
<<heterosis_mcmcem, dependson=c('heterosis_data','heterosis_model'), cache=TRUE, echo=FALSE, results='hide'>>=
current = truth
n_iter = 100
keep = data.frame(iteration = 1:n_iter, 
                  mu_phi    = rep(NA, n_iter), 
                  mu_alpha  = rep(NA, n_iter), 
                  mu_delta  = rep(NA, n_iter), 
                  mu_psi    = rep(NA,n_iter),
                  tau_phi   = rep(NA, n_iter),
                  tau_alpha = rep(NA, n_iter), 
                  tau_delta = rep(NA, n_iter),
                  tau_psi   = rep(NA,n_iter))

for (i in 1:n_iter) {
  # Run MCMC using rstan
  draws = ddply(d, .(gene), function(x) {
    s = sampling(h_m, data = c(list(n = length(x$y), count=x$y, variety = x$variety), as.list(current)))
    e = extract(s, c('phi','alpha','delta','psi'))
    data.frame(phi   = as.numeric(e$phi),
               alpha = as.numeric(e$alpha),
               delta = as.numeric(e$delta),
               psi   = as.numeric(e$psi))
  })

  current = data.frame(mu_phi = mean(draws$phi)) %>%
    mutate(mu_alpha = mean(draws$alpha),
           mu_delta = mean(draws$delta),
           mu_psi   = mean(draws$psi),
           tau_phi    = mean((draws$phi   - current$mu_phi)^2),
           tau_alpha  = mean((draws$alpha - current$mu_alpha)^2),
           tau_delta  = mean((draws$delta - current$mu_delta)^2),
           tau_psi    = mean((draws$psi   - current$mu_psi)^2))

  keep$mu_phi[i]   = current$mu_phi;   keep$tau_phi[i]   = current$tau_phi
  keep$mu_alpha[i] = current$mu_alpha; keep$tau_alpha[i] = current$tau_alpha
  keep$mu_delta[i] = current$mu_delta; keep$tau_delta[i] = current$tau_delta
  keep$mu_psi[i]   = current$mu_psi;   keep$tau_psi[i]   = current$tau_psi
}
@

<<dependson=c('normal_hierarchical_mean_em','normal_hierarchical_mean_mcem'), echo=FALSE>>=
ggplot(melt(keep, id.var=c("iteration"), variable.name="parameter"),
       aes(iteration, value)) +
  geom_point() + 
  facet_wrap(~parameter, scales="free_y", nrow=2) +
  geom_hline(data=melt(truth, variable.name='parameter'), aes(yintercept=value))
@
\end{frame}
